Profiles of Wind Direction and Speed in Turbulent Ekman Flow

### 2021-09-17 write/manuscript.tex
minor changes in Sections 3 and 4

\begin{abstract}
  The profiles of wind speed and direction in turbulent Ekman flow
  are formulated based on asymptotic theory and data from direct numerical simulation. 
  %
  The profile of the stream-wise component follows the classical viscous, logarithmic and wake scaling.
  %
  In the outer layer, the velocity component profiles can be described by an Ekman-spiral with adapted
  boundary conditions that result in a reduction of the spiral-like rotation. 
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic matching; it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  Viscous stress scales universally across the boundary layer in inner units while the total stress
  becomes universal as a function of outer height.
  %
  This implies a mixed scaling for the turbulent stress and eddy viscosity across the inner layer
  and convergence to a universal scaling as function of the outer height across the outer layer
  for increasing scale separation vide Reynolds numbers. 
\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

@@

simulation across a range of Reynolds numbers to formulate both the stream-wise and span-wise components

@@

The coordinate $Oy$ points away from the wall,  and $Oz$ points in the span-wise direction normal to $Oxy$.

@@

(i)   geostrophic wind magnitude $G=\sqrt{G_1^2+G_3^2}$ (where $\mathbf{G}=(G_1,G_2,0)$ is the geostrophic wind vector),

@@

These yield a Reynolds number $\RE$ as non-dimensional parameter governing the system.

@@

We use the geostrophic wind as velocity and the Coriolis parameter $f$ as time scale for the non-dimensional framework which implies the Rossby radius $\LR=G/f$ as length scale, such that the Reynolds number governing the problem reads as

@@

The scales used in defining $\RE_\Lambda$ loose there relevance in describing the turbulent state of the flow. The friction Reynolds number

@@

  \RE_\tau = \frac{u_\star\delta}{\nu} = \frac{\RE_\Lambda}{Z_\star^2}, 

@@

where $\delta=u_\star/f$, gives a measure for the turbulence scale separation that is comparable to other
wall-bounded flow geometries such as rotating or non-rotating channel, pipe, or Couette flow.

@@

The non-dimensionalized governing equations read as 

@@

@@

(Alternatively normalized TKE and dissipation,

@@

dependence on $\RE$ for the variation of $u_s\tar$.)

@@

The profile of the shear-aligned streamwise velocity component are well-collapsed for the Reynolds

@@

While the profiles $U^{\alpha_\star+}(z^+)$ diverge between different $\Re$ beyond $z^-=0.15$,

@@

It is noteworthy that also in the outer layer of the flow, $u_\star$--and not the magnitude the
geostrophic wind speed $G$--is the governing velocity scale for inviscid, i.e. $\RE$-independent,

@@

$S^+(z^-)$ is universal.

@@

This implies a mixed scaling for the turbulent stress, 

@@

Finally in the upper part of the outer layer, around $y^+\approx 1000$, no clear visual signature of the

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Formulations for the outer layer that take into account the rotation, and thus deviation from the channel-flow analogy
need to be matched to the framework of surface similarity, and it
occurs that a smooth transition from the inner to the outer layer is not easily achieved.
%
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when coming approaching the Ekman layer from below.
%
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to,
we believe that it is the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of geostrophic drag, $Z_\star$, and surface veering, $\alpha_\star$, with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data
    \label{fig:drag_law}}
\end{figure}
% 
\subsection{Drag-Law}
\label{sec:drag-law}
A drag-law for Ekman flow determines two key parameters of the flow as a function of Reynolds number alone.
%
These are, first, the modulus of the surface friction normalized by the geostrophic wind forcing, $u_\star$ and,
second, the direction of surface shear stress, $\alpha_\star$, also termed wind veering. 
%
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an
engineering context, and it confronts us with a situation where the most appropriate coordinate system for analysis
(namely that aligned with the surface shear stress) is a priori unknown. 
%
We compare our DNS data against a semi-empirical drag-law based on integral consideration \cite{spalart:JFM1989}
and find, as demonstrated in previous work \cite{ansorge:BM2014}, excellent agreement in the range
 $400<\mathbf{Re}<1600$, representing a factor of 16 in variation of viscosity. 
%
\par
%
We also find that the solution of the transient equation involved in estimation of $u_\star$ for a given Reynolds number
$Re_D$ is approximated reasonably by the formulation
\begin{equation}
  u_\star= 4\log(\mathbf{Re}_D))-8
\end{equation}
which quantifies the 'weak' dependence of $u_\star$ on the Reynolds number as an approximately logarithmic one, at least
for problems with a scale separation on the order that is relevant to geophysical problems ($Re_D<10^{8}$).  
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection{Profile in the outer layer}
%
The outer layer of Ekman flow is characterized by a turning of the wind velocity and the super-geostrophic maximum
that is sustained by momentum convergence at the inflection point of the velocity profile. 
The super-geostrophic maximum of streamwise velocity and a secondary minimum aloft the bulk-turbulent part of
the boundary layer are well-described by a classic Ekman spiral with adapted boundary conditions and a shift
in reference height.
%
\begin{figure}
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
  \caption{} 
\end{figure}

\subsection{Stream-wise Velocity Component} 
For the stream-wise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point.
%
\par
%
\begin{subequations}

@@

  (U,W)^{\alpha_\star+}  = (z^+,0)  

@@

\subsection{Span-wise velocity}

@@

While the universal profiles in vicinity of the wall implies that the profile be zero or at least small with respect to the stream-wise

@@

Outer scaling of the velocity profile further implies that the velocity deficity of the spanwise component $V^\alpha_star-G^\alpha_\star$

@@

  V^\alpha_\star-G^\alpha_\star = f^\alpha_\star_V(z^-).  

@@

While the stream-wise velocity follows a universal inner scaling that has acquired its universal, $Re$-independent shape for $Re_D> \mathcal{O}\left(10^3\right)$, the span-wise component that defines how the velocity vector veers when the surface is approached,

>>>>

\begin{abstract}
  The profiles of wind speed and direction in turbulent Ekman flow
  are formulated based on asymptotic theory and data from direct numerical simulation. 
  %
  The profile of the streamwise component follows the classical viscous, logarithmic and wake scaling.
  %
  In the outer layer, the velocity component profiles can be described by an Ekman-spiral with adapted
  boundary conditions that result in a reduction of the spiral-like rotation. 
  %
  The span-wise component poses a conceptual challenge to the channel-flow analogy
  in the context of asymptotic matching; it exhibits a mixed scaling in the surface layer, but follows
  outer scaling for most of the outer layer.
  %
  Viscous stress scales universally across the boundary layer in inner units while the total stress
  becomes universal as a function of outer height.
  %
  This implies a mixed scaling for the turbulent stress and eddy viscosity across the inner layer
  and convergence to a universal scaling as function of the outer height across the outer layer
  for increasing scale separation vide Reynolds numbers. 
\end{abstract}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Introduction}

@@

simulation across a range of Reynolds numbers to formulate both the streamwise and span-wise components

@@

The coordinate $Oz$ points away from the wall, and $Oy$ points in the span-wise direction normal to $Oxz$.

@@

(i)   geostrophic wind magnitude $G=\sqrt{G_1^2+G_2^2}$ (where $\mathbf{G}=(G_1,G_2,0)$ is the geostrophic wind vector),

@@

These yield a Reynolds number as non-dimensional parameter governing the system.

@@

We use the geostrophic wind as velocity and the Coriolis parameter $f$ as time scale for the non-dimensional
framework.
This implies the Rossby radius $\LR=G/f$ as length scale, such that one Reynolds number
governing the problem reads as

@@

The scales used in defining $\RE_\Lambda$ loose their relevance in describing the turbulent state of the flow.
The friction Reynolds number

@@

  \RE_\tau = \frac{u_\star\delta}{\nu} = \delta^+ = \frac{\RE_\Lambda}{Z_\star^2}, 

@@

where $\delta=u_\star/f$ and we use a superscript $'+'$ to denote normalization by inner turbulence
scales ($u_\star, \nu$).
$\RE_\tau$ is suitable to compare the turbulence scale separation also to other wall-bounded flow geometries
such as rotating or non-rotating channel, pipe, or Couette flow.

@@

The governing equations  non-dimensionalized by $G$, $f$, and $\Lambda_\mathrm{\RO}$ read as 

@@

\label{sec:scaling}

@@

(TKE and dissipation normalized as

@@

dependence on $\RE$ for the variation of $u_\star$.)

@@

The profiles of the shear-aligned streamwise velocity component are well-collapsed for the Reynolds

@@

While the profiles $U^{\alpha_\star+}(z^+)$ diverge between different $\RE$ beyond $z^-=0.15$,

@@

It is noteworthy that also in the outer layer of the flow, $u_\star$ (not the magnitude the
geostrophic wind speed $G$) is the governing velocity scale for inviscid, i.e. $\RE$-independent,

@@

$S^+(z^-)$ is universal; a discrepancy in the inner layer does not occur as the total stress
is approximately constant in the viscous and buffer layer, and a rescaling of the height would have
no effect there; above, outer scaling is appropriate for the well-established dynamics 
in the overlap region of inner and outer layer.

@@

This, however, implies a mixed scaling for the turbulent stress, 

@@

In the upper part of the outer layer, around $y^+\approx 1000$, no clearly visible signature of the

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
\label{sec:profiles} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We now turn to the formulation of a general velocity profile that is fully determined by the only
parameter of the idealized problem, namely a Reynolds number representing the scale separation or
geometric size of the problem.
%
This precludes a drag law with wherewith we begin this section (\ref{sec:drag-law}).
Based on the scaling arguments put forward in Sec.~\ref{sec:scaling}, we then develop
a formulation of the wind vector in the Ekman layer (Sec~\ref{sec:ekman}) and separately
for the streamwise (Sec.~\ref{sec:streamwise}) and spanwise components (Sec.~\ref{sec:spanwise})
in the overlap and inner regions. 
%

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of geostrophic drag, $Z_\star$, and surface veering, $\alpha_\star$, with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data
    \label{fig:drag_law}}
\end{figure}
% 
\subsection{Drag-Law}
\label{sec:drag-law}
A drag-law for Ekman flow determines two key parameters of the flow as a function of Reynolds number alone.
%
These are, first, the modulus of the surface friction normalized by the geostrophic wind forcing, $u_\star$ and,
second, the direction of surface shear stress, $\alpha_\star$, also termed wind veering. 
%
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an
engineering context, and it confronts us with a situation where the most appropriate coordinate system for analysis
(namely that aligned with the surface shear stress) is a priori unknown. 
%
We compare our DNS data against a semi-empirical drag-law based on integral consideration \cite{spalart:JFM1989}
and find, as demonstrated in previous work \cite{ansorge:BM2014}, excellent agreement in the range
 $400<\mathbf{Re}<1600$, representing a factor of 16 in variation of viscosity. 
%
\par
%
We also find that the solution of the transient equation involved in estimation of $u_\star$ for a given Reynolds number
$Re_D$ is approximated reasonably by the formulation
\begin{equation}
  u_\star= 4\log(\mathbf{Re}_D))-8
\end{equation}
which quantifies the 'weak' dependence of $u_\star$ on the Reynolds number as an approximately logarithmic one, at least
for problems with a scale separation on the order that is relevant to geophysical problems ($Re_D<10^{8}$).  
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection{Profile in the Ekman layer}
\label{sec:ekman}
%
Formulations for the outer layer that take into account the rotation
(and thus deviation from the channel-flow analogy)
need to be matched to the framework of surface similarity.
%
A smooth transition from the inner layer to the Ekman layer, where the wind is characterized by a turning
of its mean direction, is not easily achieved. 
%
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when coming approaching the Ekman layer from below.
%
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to,
we understand this as the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.
%
The outer layer of Ekman flow is characterized by a turning of the wind velocity and the super-geostrophic maximum
that is sustained by momentum convergence at the inflection point of the velocity profile.
%
The super-geostrophic maximum of streamwise velocity and a secondary minimum aloft the bulk-turbulent part of
the boundary layer are well-described by a classic Ekman spiral with adapted boundary conditions and a shift
in reference height.
%
\begin{figure}
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
  \caption{} 
\end{figure}

\subsection{Streamwise Velocity Component} 
For the streamwise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point.
%
\par
%
\begin{subequations}

@@

  (U,V)^{\alpha_\star+}  = (z^+,0)  

@@

\subsection{Spanwise velocity component}

@@

While the universal profiles in vicinity of the wall implies that the profile be zero or at least small with respect to the streamwise

@@

Outer scaling of the velocity profile further implies that the velocity deficity of the spanwise component $V^{\alpha_\star}-G^{\alpha_\star}$

@@

  V^{\alpha_\star}-G^{\alpha_\star} = f^{\alpha_\star}_V(z^-).  

@@

While the streamwise velocity follows a universal inner scaling that has acquired its universal, $Re$-independent shape for $Re_D> \mathcal{O}\left(10^3\right)$, the span-wise component that defines how the velocity vector veers when the surface is approached,
